\LARGE \bf Fast Approximate 2D Ray Casting for Accelerated Localization

### 2017-02-28 paper/main.tex
another major revision

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphicx}
\usepackage{tabu}
\usepackage{float}
% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{shapes.geometric, arrows}

@@

\newcommand{\imgFrame}[1]{\begin{center}\frame{\includegraphics[width=\columnwidth]{{#1}}}\end{center}}

@@

\title{\LARGE \bf
Compressed Directional Distance Transform for Fast 2D Ray Casting
}

@@

\skcomment{}{I would consider putting the word ``localization'' in the title.}

@@

While particle filters provide a robust and general framework for solving the localization problem, they can be computationally expensive due to both the number of particles which must be maintained and the evaluation of the sensor model. In robots with range sensors such as LiDAR or Sonar, ray casting is frequently used to compare sensor readings with the ground truth distance between the hypothesis pose and obstacles in a map. Ray casting itself is a complex operation, and reweighting any single particle may require tens of ray casts. Many effective particle filters maintain thousands of particles, updating each particle tens of times per second. As a result, millions of ray cast operations may be resolved per second, posing a significant computational challenge for resource constrained systems.

@@

We propose an analogous data structure to the three-dimensional LUT called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries. The contributions of this work are the following. Firstly, we develop a problem formulation and provide details of our method. Secondly, we implement the proposed algorithm and compare its performance to alternate two-dimensional ray casting methods. Finally, we discuss potential extensions to the CDDT algorithm which could be the basis of future research. We observe two orders of magnitude improvement in memory requirements, with little sacrifice in computation time, when compared to the lookup table methods. Additionally, we observe a large speedup when compared to the other ray casting methods considered with similar memory requirements.

@@

Due to varied handling of edge cases and ambiguities in the problem of determining distance between any point and the nearest obstacle in a particular direction, most 2D ray casting algorithms do not provide exactly consistent results for every query. However, on average error between results from each of the methods discussed is small.

Bresenham's line algorithm \cite{bresenham} incrementally determines the set of pixels that approximate the trajectory of query ray starting from the query point $(x,y)$ and progressing in the $\theta$ direction one pixel at a time. The algorithm terminates once the nearest occupied pixel is discovered, and the euclidean distance between that occupied pixel and the query $(x,y)$ is reported. This algorithm is widely implemented in particle filters due to its simplicity and ability to operate on a dynamic map. The primary disadvantage is that it is slow, potentially requiring hundreds of memory accesses for a single ray cast. While actual performance is highly environment dependent, Bresenham’s Line algorithm is linear in map size in the worst case.

Similar to Bresenham's Line algorithm, ray marching \cite{raymarching} checks points along the line radiating outwards from the query point until an obstacle is encountered. The primary difference is that ray marching makes larger steps along the query ray, thereby avoiding unnecessary memory reads. Beginning at the query point $(x,y)$ the ray marching algorithm proceeds in the theta direction, stepping along the line by the minimum distance between each query point and the nearest obstacle. The algorithm terminates when the query point coincides with an obstacle in the map. A precomputed euclidean distance transform of the occupancy map provides the distance between each query point and the nearest obstacle in any direction.

@@

Ray marching is on average much faster than Bresenham’s line, but edge cases exist in which the performance is nearly equivalent. As noted in \cite{acceleratedraymarching} the traversal speed of rays rapidly decreases as sampled positions approach obstacles. For this reason, rays which travel parallel to walls progress very slowly as compared to those passing through open areas. Thus the performance of ray marching exhibits a long tail distribution as seen in (Fig. \ref{violin:basement:all}, \ref{violin:basement:particle}) which negatively impacts average runtime as compared to the median, and can be problematic for near real time algorithms.

As previously described, a common acceleration technique for two dimensional ray casting is to precompute the ray distances for every state in a discrete grid and store the results in a three dimensional LUT. Any ray casting technique may be used to compute the table. In the authors’ implementation ray casting is used to populate the LUT since it has low initialization time and is significantly faster than Bresenham’s Line. Theoretically this approach has constant query runtime, though actual performance is dependent on access patterns as CPU caching effects are significant in practice.

State space discretization implies approximate results, since intermediate states must be rounded onto the discrete grid. While the effect of rounding query position, it can be significant for $\theta$. As the ray moves further away from the query point along the theta direction, angular discretization error accumulates. For queries $(x,y,\theta)$ discretized into $(x_d,y_d,\theta_d)$, the distance between the end of the ray $(x_d,y_d,Θ_d)$ and its projection onto the line implied by $(x,y,\theta)$ becomes large as the length of the ray increases. Although the error induced by discretization may be unacceptable for computer graphics applications, it is generally acceptable for probabilistic algorithms such as the particle filter since error is expected and directly modeled.

Rather than simply rounding to the nearest grid state, one method to improve accuracy of queries which lie between discrete states is to query the neighboring states and interpolate between them. Since we are evaluating the performance of these methods for use in a particle filter, we do not perform interpolation since the increased computation reduces the number of particles that may be maintained in real time, and therefore any increase in accuracy with respect to the stored map is effectively negated.

@@

We define the problem of ray casting in occupancy grids as follows. We assume a known occupancy grid map in which occupied cells have value one and unoccupied cells have value zero. Given a query pose $(x,y,\theta)_{query}$ in map space, the ray cast operation finds the nearest occupied pixel $(x,y)_{collide}$ which lies on the ray starting at the position $(x,y)_{query}$ pointing in the $\theta_{query}$ direction. The ray cast operation returns the euclidean distance $d_{ray}$ between $(x,y)_{query}$ and $(x,y)_{collide}$.

@@

We refer to $(x,y,\theta)_{query}$ as the query pose, and $(x,y)_{query}$ as the query point. Additionally, denote the discretized query pose as $\lfloor(x,y,\theta)_{query}\rceil$. A $\theta$ slice through the LUT refers to all table entries for that particular $\lfloor\theta\rceil$.

@@

Although the three dimensional table used to store precomputed ray cast solutions for a discrete state space is inherently large, it is highly compressible. This is most apparent in the cardinal directions, in which adjacent values along a particular dimension of the table increase by exactly one unit of distance for unobstructed positions as in Fig. \ref{cardinal_example}. Our data structure is designed to compress this redundancy while still allowing for fast queries in near constant time. We accomplish this though through what we refer to as a Compressed Directional Distance Transform (CDDT) described here.

@@

In contrast, to compute the DDT, the obstacles in the map are rotated about the origin by $-\lfloor\theta\rceil$ and ray casting is implicitly performed in the $\theta=0$ direction to populate the table. Then, to index the table for a query $(x,y,\theta)_{query}$ one discretizes to $\lfloor(x,y,\theta)_{query}\rceil$, rotates $\lfloor(x,y)_{query}\rceil$ about the origin, and uses the rotated coordinates $\lfloor(x,y,\theta)_{query}\rceil_{rot}$ to index into the DDT. Thus, the same operation is computed in both the DDT and the LUT, but while one changes the ray cast direction to populate the LUT, one rotates scene geometry and ray casts in a constant direction to populate the DDT. While it is true that the rotation of the scene geometry introduces small errors as compared to ray casting for each cell, as previously discussed small errors do not significantly impact particle filter localization performance due to the use of a probabilistic error model and are dominated by errors inherently present in distance sensors.

@@

The conversion from a DDT to a CDDT slice is performed by storing the $x$ coordinate of each zero point in the coordinate space of the DDT for every discrete row. At query time, the sawtooth function (which is the distance to the nearest obstacle in the query direction) encoded in the DDT may be quickly recovered by performing binary search for nearest zero point in the correct row of the CDDT slice. We refer to the list of zero points for a single row as a CDDT bin. Similar to the LUT, the full CDDT is defined as every CDDT slice for all discrete values of $\theta$.

For performance, it is not necessary to compute the full DDT, but rather the CDDT may be directly constructed by projecting each obstacle into the coordinate space of the DDT for every $\lfloor\theta\rceil$ and sorting its $x$ coordinate in the CDDT bin corresponding to its $y$ coordinate and rotation. After all geometry has been projected into the CDDT, each bin is sorted to facilitate later indexing. Not only does the direct construction of the CDDT greatly reduce the amount of memory required to store the LUT, it also reduces data structure precomputation time since ray casting is not necessary.

[TODO add an example of scene rotation, saw tooth wave, and zero point projection]

@@

@@

{\color{red}DONE WITH SECOND EDITING PASS UP TO THIS POINT}

Once the CDDT is constructed, querying the data structure to resolve ray casts involves

\begin{enumerate}
    \item Projecting the query point into the coordinate space of the slice which corresponds to the discretized theta
    \item Finding the nearest zero point which has a y-coordinate greater than or equal to the query point in the projected coordinate space
    \item Return the distance between the query point and the nearest zero point in the projected coordinate space, which is simply the difference between y-coordinates
\end{enumerate}

Binary search is theoretically optimal for finding zero points near to query states, in our implementation we conditionally use linear search when the number of zero points is below a certain constant (we use 64) as it has a superior memory access pattern in such cases.
While conceptually simple, implementing the CDDT data structure construction and traversal routines requires careful consideration in order to capture all edge cases and to minimize unnecessary computation. In this sense, it is more complex to implement than the alternatives considered, however there are many opportunities for optimization which yield real-world speed up. To ease this burden, we provide our implementation as well as Python wrappers with an Apache 2.0 license.

\skcomment{}{You can put all the text after this point into one subsection, which you can call ``Further optimizations'' or something like that.}

\subsection{Further Optimizations}

A side effect of extracting the zero points from each column of the DDT is the introduction of rotational symmetry. For any given theta, the zero points accumulated by projecting scene geometry into the requisite coordinate space are mirror images of those accumulated by projecting scene geometry into the theta+pi direction. Therefore, one need only compute the CDDT for the range of discrete thetas from $[0,\pi)$ and the DDT for the range $[0,2\pi)$ may be inferred, resulting in factor of two further reduction in memory usage.

In addition to the aforementioned reduction in memory usage, rotational symmetry may be exploited in scenarios where ray casts are performed radially around a single point. While traversing the data structure to resolve a ray cast in a particular $(x,y,\theta)$, the ray cast for $(x,y,\theta+\pi)$ may be resolved with a single additional memory read. Once the index i of the nearest zero point in the theta direction $z_{\theta}$ is discovered, the index of the nearest zero point in the opposite direction is simply i-1. For example, in robots with laser scanners sweeping angles larger than 180°, this symmetry can be used to reduce the number of data structure traversals required to compute the sensor model by up to a factor of two.

\subsection{Incremental Update}

While the authors did not implement the necessary changes, we conjecture that the CDDT algorithm could be modified to allow incremental updates to the map. In contrast, any change to the map would introduce inconsistencies in a precomputed LUT requiring a full re-computation of the data structure. Ray casting every state in a three dimensional grid is highly computational, ruling out the possibility of using a precomputed LUT with dynamic maps such as in SLAM algorithms. Similarly, ray marching operates on a euclidean distance transform which cannot be easily updated in an incremental fashion. While Bresenham’s line algorithm is the slowest option evaluated in terms of ray cast performance, it has the benefit that it operates directly on the occupancy grid map and can therefore be used with dynamic maps. For this reason, Bresenham’s line is one of the most commonly used algorithms for ray casting in occupancy grids.

To allow incremental update of the CDDT, the sorted vector data structure used in our implementation should be replaced with an alternative which allows for $O(logn)$ insert and delete, such as a randomized skip list. To aid in node deletion, pointers could be maintained from each grid element to the associated leaf node in the randomized skip lists with a constant factor increase in memory usage.

In this case, the cost of inserting or deleting an occupied grid cell would be \skcomment{$O(theta\_discretization*\log(number of lut bin elements))$.}{I think this is your plan already, but just to note: Please make sure to use symbols for these and introduce what the symbols mean. } The number of elements in each lut bin tends to be small and is bounded, so this becomes $O(theta\_discretization)$ which is likely not prohibitive for real-time performance.

Skip list traversal is likely to be less efficient than binary search on sorted vectors due to worse cache characteristics. However, the authors expect it to still be significantly faster than performing Bresenham’s line, and therefore recommend that this modification be the subject of future research.

\subsection{Pruning}

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. \skcomment{Consider a 3x3 block of obstacles in an otherwise empty map.}{I think this deserves a figure, just to allow the reader to quickly read your paper.} The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. This can result in significant memory usage and data structure construction time reductions in dense maps where the number of obstacles adjacent to empty cells is small compared to the total number of obstacles. Prior to performing ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles which may be removed by the morphological operation.

\begin{figure}[h]
\img{slice0_pcddt_basement.png}
\caption{The same slice of the DDT as \ref{cddt:slice0}. Notice that the number of elements projected into each CDDT bin is less peaky since the pruning operation removes most elements of long lines of obstacles. Also note the higher compression factor of 415 for this slice, increased from 267 in the unpruned CDDT.}
\label{pcddt:slice0}
\end{figure}

\skcomment{Additionally, consider a line of obstacles aligned along the Y-axis.}{While you are at it, put a figure for this one too? } Every element in this line will be projected into a single list of zero points for the theta=0 slice of the CDDT. However, the middle elements of the line will never result in a collision. Any ray cast from adjacent states in the $\theta=0$ direction will return early in the occupancy grid check, and any ray cast from outside the wall will collide with one of the edge-most obstacles in that line. Therefore in the theta=0 slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary theta not aligned with an axis. Rather than attempting to analytically determine which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state (as in the LUT data structure) and discard any zero point which is unused.

Pruning does also increase precomputation time rendering it incompatible with incremental update, however the reduction of memory usage is worthwhile for static maps. In addition to memory reduction, we find that pruning slightly improves runtime performance, likely as a result of improved caching characteristics.

\skcomment{}{Do you think you can have a subsection called ``analysis'' which has a few propositions/theorems that say things like Memory requirement is $O(NM)$ and running time is $O(1)$? This may be hard given the deadline, but osmething to think about. }

\section{EXPERIMENTS}

@@

@@

\subsection{Synthetic Benchmarks}

>>>>

% The following packages can be found on http:\\www.ctan.org
\usepackage{graphicx}
\usepackage{tabu}
\usepackage{float}
% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{shapes.geometric, arrows}
\usepackage{algorithm}
% \usepackage[]{algorithm2e}
\usepackage{algorithmic}

@@

\newcommand{\imgFrame}[1]{\begin{center}\frame{\includegraphics[width=\columnwidth]{{#1}}}\end{center}}
\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}

@@

\title{\LARGE \bf
Fast Approximate 2D Ray Casting for Accelerated Localization
% Compressed Directional Distance Transform for Fast 2D Ray Casting
}

@@

@@

While particle filters provide a robust and general framework for solving the localization problem, they can be computationally expensive due to both the number of particles which must be maintained and the evaluation of the sensor model. In robots with range sensors such as LiDAR or Sonar, ray casting is frequently used to compare sensor readings with the ground truth distance between the hypothesis pose and obstacles in a map. Ray casting itself is a complex operation, and a single evaluation of the sensor model may require tens of ray casts. Many effective particle filters maintain thousands of particles, updating each particle tens of times per second. As a result, millions of ray cast operations may be resolved per second, posing a significant computational challenge for resource constrained systems.

@@

We propose an analogous data structure to the three-dimensional LUT called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries. The contributions of this work are the following. Firstly, we develop a problem formulation and provide details and analysis of our method. Secondly, we implement the proposed algorithm and compare its performance to alternate two-dimensional ray casting methods. Finally, we discuss potential extensions to the CDDT algorithm which could be the basis of future research. We observe two orders of magnitude improvement in memory requirements, with little sacrifice in computation time, when compared to the lookup table methods. Additionally, we observe a large speedup when compared to the other ray casting methods considered with similar memory requirements.

@@

Bresenham's line algorithm \cite{bresenham} incrementally determines the set of pixels that approximate the trajectory of query ray starting from the query point $(x,y)_{query}$ and progressing in the $\theta_{query}$ direction one pixel at a time. The algorithm terminates once the nearest occupied pixel is discovered, and the euclidean distance between that occupied pixel and $(x,y)_{query}$ is reported. This algorithm is widely implemented in particle filters due to its simplicity and ability to operate on a dynamic map. The primary disadvantage is that it is slow, potentially requiring hundreds of memory accesses for a single ray cast. While actual performance is highly environment dependent, Bresenham’s Line algorithm is linear in map size in the worst case.

Similar to Bresenham's Line algorithm, ray marching \cite{raymarching} checks points along the line radiating outwards from the query point until an obstacle is encountered. The primary difference is that ray marching makes larger steps along the query ray, thereby avoiding unnecessary memory reads. Beginning at $(x,y)_{query}$ the ray marching algorithm proceeds in the $\theta_{query}$ direction, stepping along the line by the minimum distance between each query point and the nearest obstacle. The algorithm terminates when the query point coincides with an obstacle in the map. A precomputed euclidean distance transform of the occupancy map provides the distance between each query point and the nearest obstacle in any direction.

@@

Ray marching is on average much faster than Bresenham’s line, but edge cases exist in which the theoretically asymptotic runtime is equivalent. As noted in \cite{acceleratedraymarching} the traversal speed of rays rapidly decreases as sampled positions approach obstacles. For this reason, rays which travel parallel to walls progress very slowly as compared to those passing through open areas. Thus the performance of ray marching exhibits a long tail distribution as seen in (Fig. \ref{violin:basement:all}, \ref{violin:basement:particle}) which negatively impacts average runtime as compared to the median, and can be problematic for near real time algorithms.

As previously described, a common acceleration technique for two dimensional ray casting is to precompute the ray distances for every state in a discrete grid and store the results in a three dimensional LUT. Any ray casting technique may be used to compute the table. In the authors’ implementation ray casting is used to populate the LUT since it has low initialization time and is significantly faster than Bresenham’s Line. Theoretically the LUT approach has constant query runtime, though actual performance is dependent on access patterns as CPU caching effects are significant in practice.

State space discretization implies approximate results, since intermediate states must be rounded onto the discrete grid. While the effect of rounding query position is small, rounding $\theta_{query}$ may have significant effects. As the ray moves further away from the query point along the theta direction, angular discretization error accumulates. For queries $(x,y,\theta)_{query}$ discretized into $\lfloor(x,y,\theta)\rceil$, the distance between the end of the ray $\lfloor(x,y,\theta)\rceil$ and its projection onto the line implied by $(x,y,\theta)_{query}$ becomes large as the length of the ray increases. Although the error induced by discretization may be unacceptable for computer graphics applications, it is generally acceptable for probabilistic algorithms such as the particle filter since error is expected and directly modeled.

More generally, due to varied handling of edge cases and ambiguities in the problem of determining distance between any point and the nearest obstacle in a particular direction, most 2D ray casting algorithms do not provide exactly consistent results for every query. However, on average error between results from each of the methods discussed is small. Any error that does exist is dominated by error inherent in distance sensors.

Rather than simply rounding to the nearest grid state, one may improve accuracy of queries which lie between discrete states by querying the neighboring states and interpolating the results. Since we are evaluating the performance of these methods for use in a particle filter, we do not perform interpolation since the increased computation reduces the number of particles that may be maintained in real time, thereby effectively negating any potential increase in accuracy with respect to the stored map.

@@

We define the problem of ray casting in occupancy grids as follows. We assume a known occupancy grid map in which occupied cells have value 1, and unoccupied cells have value 0. Given a query pose $(x,y,\theta)_{query}$ in map space, the ray cast operation finds the nearest occupied pixel $(x,y)_{collide}$ which lies on the ray starting at the position $(x,y)_{query}$ pointing in the $\theta_{query}$ direction, and returns the euclidean distance $d_{ray}$ between $(x,y)_{query}$ and $(x,y)_{collide}$.

@@

We refer to $(x,y,\theta)_{query}$ as the query pose, and $(x,y)_{query}$ as the query point. We denote the discretized query pose as $\lfloor(x,y,\theta)_{query}\rceil$. A $\theta$ slice through the LUT refers to all table entries for that particular $\lfloor\theta\rceil$. The number of discrete $\lfloor\theta\rceil$ values is denoted $\theta_{discretization}$. Fig. \ref{coordinate_system} demonstrates our chosen coordinate system.

@@

Although the three dimensional table used to store precomputed ray cast solutions for a discrete state space is inherently large, it is highly compressible. This is most apparent in the cardinal directions, in which adjacent values along a particular dimension of the table increase by exactly one unit of distance for unobstructed positions as in Fig. \ref{cardinal_example}. Our data structure is designed to compress this redundancy while still allowing for fast queries in near constant time. We accomplish this though through what we refer to as a Compressed Directional Distance Transform described here.

@@

\begin{figure}[h]
\img{ddt_lut_slice_comparison_map.png}
\caption{Comparison between a DDT and a LUT slice for the same value of theta. Each row in the DDT is characterized by a sawtooth function.}
\label{ddt_lut_comparison}
\end{figure}

In contrast, to compute the DDT, the obstacles in the map are rotated about the origin by $-\lfloor\theta\rceil$ and ray casting is implicitly performed in the $\theta=0$ direction, as demonstrated by Fig. \ref{ddt_lut_comparison}. Then, to index the table for a query $(x,y,\theta)_{query}$ one discretizes to $\lfloor(x,y,\theta)_{query}\rceil$, rotates $\lfloor(x,y)_{query}\rceil$ about the origin, and uses the rotated coordinates $\lfloor(x,y,\theta)_{query}\rceil_{rot}$ to (implicitly) index into the DDT. Thus, roughly the same operation is computed in both the DDT and the LUT, but while one changes the ray cast direction to populate the LUT, one rotates scene geometry and ray casts in a constant direction to populate the DDT.

We define $P_{DDT_{\theta}}$ as the three by three transformation matrix which projects vectors into the coordinate space of the DDT in the $\theta$ direction. While it is true that the transformation of the scene geometry introduces small errors as compared to ray casting for each cell, as previously discussed, small errors do not significantly impact particle filter localization performance.

@@

\begin{figure}[h]
\img{sawtooth_to_cddt.png}
\caption{Demonstration of compression from a sawtooth function to a list of zero points. Consecutively valued zero points exist where walls are greater than one pixel thick.}
\label{sawtooth_to_cddt}
\end{figure}

The conversion from a DDT to a CDDT slice is performed by storing the $x$ coordinate of each zero point in the coordinate space of the DDT for every discrete row as shown in Fig. \ref{sawtooth_to_cddt}. At query time, the sawtooth function (i.e. exactly the distance to the nearest obstacle in the query direction) encoded in the DDT may be quickly recovered by performing binary search for nearest zero point in the correct row of the CDDT slice. We refer to the list of zero points for a single row as a CDDT bin. Similar to the LUT, the full CDDT is defined as every CDDT slice for all discrete values of $\theta$.

For performance, it is not necessary to compute the full DDT, but rather the CDDT may be directly constructed by projecting each obstacle into the coordinate space of the DDT for every $\lfloor\theta\rceil$ and sorting its $x$ coordinate in the CDDT bin corresponding to its $y$ coordinate and $\lfloor\theta\rceil$. After all geometry has been projected into the CDDT, each bin is sorted to facilitate later indexing. Not only does the direct construction of the CDDT greatly reduce the amount of memory required to store the LUT, it also reduces data structure precomputation time since ray casting is not necessary.

While conceptually simple, implementing the CDDT data structure construction and traversal routines requires careful consideration in order to capture all edge cases and to minimize unnecessary computation. In this sense, it is more complex to implement than the alternatives considered, however there are many opportunities for optimization which yield real-world speed up. To ease this burden, we provide our implementation as well as Python wrappers with an Apache 2.0 license.

@@

\begin{center}
\line(1,0){240}
\end{center}
\begin{algorithmic}
% \STATE {\color{gray}\# Finds all occupied pixels bordering unoccupied pixels}
\STATE $edge\_map\gets map-morphological\_erosion(map)$
\STATE Initialize $\theta_{discretization}$ empty CDDT slices
\FOR {$\theta \in \{\lfloor\theta\rceil\}$} 
        \FOR{each occupied pixel $(x,y)\in edge\_map$}
            \STATE $(x,y)_{DDT_{\theta}} = P_{DDT_{\theta}}*\begin{pmatrix}
                        x \\
                        y \\
                        1
             \end{pmatrix}$
             \FOR {each CDDT bin overlapping with $y_{DDT_{\theta}}$}
                \STATE bin.append($x_{DDT_{\theta}}$)
             \ENDFOR
        \ENDFOR
        \FOR{each CDDT bin}
            \STATE bin = sort(bin)
        \ENDFOR
\ENDFOR
\end{algorithmic}
\begin{center}
\line(1,0){240}
\end{center}

@@

\begin{center}
\line(1,0){240}
\end{center}
\begin{algorithmic}
\STATE \textbf{function ray\_cast($(x,y,\theta)_{q})$}
\INDSTATE[1] $(x,y)_{DDT_{\theta_{q}}} = P_{DDT_{\theta_{q}}}*\begin{pmatrix}
                        x_{q} \\
                        y_{q} \\
                        1
             \end{pmatrix}$
\INDSTATE[1] bin $\gets$ zero points in row $y_{DDT_{\theta_{q}}}$ of CDDT slice $\theta_{q}$
\INDSTATE[1] $x_{collide}$ = smallest element $x_{collide} > x_{DDT_{\theta_q}}\in$ bin
% find nearest greate element in CDDT bin correspond to y_ddt
\INDSTATE[1] \textbf{return} abs($x_{DDT_{\theta_q}} - x_{collide}$)
\STATE \textbf{end function}\end{algorithmic}
\begin{center}
\line(1,0){240}
\end{center}

In our implementation we switch between linear and binary search to find $x_{collide}$ in a given CDDT bin depending on the number of zero points in that bin. While binary search is theoretically optimal for sorted arrays, in practice linear search has a superior memory access pattern and is therefore faster in small searches.

\subsection{Further Optimizations} \label{optimizations}

A side effect of extracting zero points from each row of the DDT is the introduction of rotational symmetry. Not only can a row of the DDT in the $\theta$ direction be reconstructed from the zero points, but also a row of the DDT in the $\theta+\pi$ direction as in Fig. \ref{bidirectional_zero_points}. Therefore, one only need compute CDDT slices for $0\leq\lfloor\theta\rceil<\pi$ and the DDT for $0\leq\lfloor\theta\rceil<2\pi$ may be inferred, resulting in a factor of further two reduction in memory usage.

\begin{figure}[h]
\img{bidirectional_zero_points.png}
\caption{Demonstration of reconstructing rows of two DDT slices from a single set of zero points.}
\label{bidirectional_zero_points}
\end{figure}

In addition to the aforementioned reduction in memory usage, rotational symmetry may be exploited in scenarios where ray casts are performed radially around a single point. While traversing the data structure to resolve a ray cast query $(x,y,\theta)$, the ray cast for $(x,y,\theta+\pi)$ may be resolved with a single additional memory read. Once a search algorithm is used to discover the index $i$ of $x_{collide_{\theta}}$ in the CDDT, the index of $x_{collide_{\theta+\pi}}$ is simply $i-1$ as in Fig. \ref{bidirectional_zero_points}. For example, this symmetry can be used in robots with laser scanners sweeping angles larger than $180^{\circ}$ to reduce the number of data structure traversals required to compute the sensor model by up to a factor of two.

% \subsection{Incremental Update}

While the authors did not implement the necessary changes, we conjecture that the CDDT algorithm could be modified to allow incremental updates to the map. In both ray marching and the LUT approach, any change in the map introduces inconsistencies to the acceleration data structure requiring a full re-initialization. As demonstrated by our experiments, initializing the three dimensional LUT is a highly computational process. Similarly, computing a euclidean distance transform for ray marching can be expensive. Therefore neither ray casting nor the LUT can be used with dynamic maps such as in SLAM algorithms. Despite being the slowest option evaluated, Bresenham's Line algorithm is one of the most commonly used techniques since it operates directly on the grid map without an acceleration data structure.

To allow incremental update of the CDDT, the sorted vector data structure used in our implementation to store zero points should be replaced with an alternative which allows for $O(\log{n})$ insertion and deletion, such as a randomized skip list. To aid in zero point deletion, pointers could be maintained correlating each cell in the occupancy grid with its associated zero points for each CDDT slice with a constant factor increase in memory usage. In this case, the cost of toggling the state of a cell in the occupancy grid would be $O(\theta_{discretization}\log{n})$ where $n$ is the number of elements in each associated CDDT bin. For most occupancy grid maps of real environments, the value of $n$ is small, and in any case is bounded by the size of the map, so $\log{n}$ is bounded by a small constant factor. Therefore, the cost of update with becomes $O(\theta_{discretization})$ which is likely not prohibitive for real-time performance in dynamic maps.

Skip list traversal is likely to be less efficient than binary search on sorted vectors due to worse cache characteristics. However, the authors expect it to still be faster than performing Bresenham’s Line algorithm, and therefore recommend that this modification be the subject of future research.

% \subsection{Pruning}
\begin{figure}[h]
\begin{center}\includegraphics[width=6.5cm]{edge_map.png}\end{center}
\caption{Example map left, morphological edge map right.}
\label{edge_map}
\end{figure}

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. Consider a 3x3 block of obstacles in an otherwise empty map as in Fig. \ref{edge_map}. The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. Prior to performing ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles removed during the morphological operation. In dense maps, this results in a significant reduction in both memory usage and construction time.

% \begin{figure}[h]
% \img{slice0_pcddt_basement.png}
% \caption{A slice of the PCDDT visualized as the implied DDT. Blue line demonstrates the num}
% \label{pcddt:slice0}
% \end{figure}

Additionally, consider a line of obstacles aligned along the X-axis. Every element in this line will be projected into a single zero point bin in the $\theta=0$ CDDT slice. However, the middle elements of the line will never result in a collision. Any ray cast from points on the line of obstacles will return early in the occupancy grid check, and any ray cast from non-overlapping points co-linear in the $\theta$ or $\theta+\pi$ directions will intersect one of the edge-most obstacles in the line. Therefore in the $\theta=0$ CDDT slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary $\lfloor\theta\rceil$ not aligned with an axis. Rather than analytically determining which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state, discarding any unused zero point.

Pruning does increase pre-computation time, rendering it incompatible with incremental update. However, the reduction of memory usage is worthwhile for static maps. In addition to memory reduction, we find that pruning slightly improves runtime performance, likely as a result of improved caching characteristics and the reduced number of zero points.

\section{Analysis}

% \skcomment{}{Do you think you can have a subsection called ``analysis'' which has a few propositions/theorems that say things like Memory requirement is $O(NM)$ and running time is $O(1)$? This may be hard given the deadline, but osmething to think about. }

The memory usage of the CDDT data structure is $O(n*\theta_{discretization}+|map|)$ where $n$ is the number of occupied pixels in the edge map. The occupancy grid map must be stored to check for overlaps between $(x,y)_{query}$ and obstacles prior to searching CDDT bins. For each edge pixel, a total of $\theta_{discretization}$ float values are stored in the CDDT bins. Due to the fact that we must sort each bin after CDDT construction, pre-computation time is at worst $O(n*\theta_{discretization} + S^2*\theta_{discretization}*log(S))$ for the same definition of n, where $S=norm(\text{map width},\text{map height})$. In practice it is closer to $O(n*\theta_{discretization} + S*\theta_{discretization})$ since each CDDT has a small number of elements on average, as evidenced by the high demonstrated compression ratio.

The pruning operation described in \ref{optimizations} reduces memory requirement, with a computational cost of $O(|map|*\theta_{discretization}*O(calc\_range)_{CDDT})$. The precise impact of pruning is scene dependent and difficult to analyze in the general case.

The ray cast procedure has three general steps: projection into CDDT coordinate space, the search for nearby zero points, and the computation of distance given the nearest zero point. The first and last steps are simple arithmetic, and therefore are theoretically constant time. The second step involves searching for a value in a sorted array of zero points. As previously discussed in \ref{optimizations}, the number of zero points in each CDDT bin tends to be small and is bounded in map size. Thus, at worst the search operation requires $\log{(norm(\text{map width},\text{map height}))}$ which is a constant value for a fixed map. Therefore, for a given map, our algorithm provides constant time performance.

\section{Experiments}

@@

\begin{figure}[h!]
\begin{center}
\begin{tabular}{ | m{8em} | m{2.3cm}| m{1.5cm} | } 
\hline
 \multicolumn{3}{|c|}{Basement Map, $\theta_{discretization}$: 108} \\
 \hline
 Method & Memory Usage & Init. Time \\
 \hline
 Bresenham's Line & 1.37 MB & 0.006 sec  \\
 Ray Marching & 5.49 MB & 0.16 sec  \\
 CDDT & 6.34 MB & 0.067 sec  \\
 PCDDT & 4.07 MB & 2.2 sec  \\
 Lookup Table & 296.63 MB & 15.3 sec  \\
\hline \hline
% \end{center}
% % \caption{The construction time and memory footprint of each method for the Basement map.}
% \label{table:basement:init}
% \end{figure}
% \begin{figure}[ht]
% \begin{center}
% \begin{tabular}{ | m{8em} | m{2.3cm}| m{1.5cm} | } 
 \multicolumn{3}{|c|}{Synthetic Map, $\theta_{discretization}$: 108} \\
 \hline
 Method & Memory Usage & Init. Time \\
 \hline
 Bresenham's Line & 1 MB & 0.004 sec  \\
 Ray Marching  & 4 MB & 0.13 sec  \\
 CDDT & 2.71 MB & 0.03 sec  \\
 PCDDT & 1.66 MB & 0.74 sec  \\
 Lookup Table & 216 MB & 9.1 sec  \\
\hline
\end{tabular}
\end{center}
\caption{Construction time and memory footprint of each method.}
\label{table:synthetic:init}
\end{figure}

@@

We evaluate algorithm performance in two synthetic benchmarks, using two different maps. The so called Synthetic map was created with photoshop, whereas the basement map was created via a SLAM algorithm on the RACECAR platform {\color{red} [cite this?]}. The first benchmark computes a ray cast for each point in a uniformly spaced grid over the three dimensional state space. The second benchmark performs a large number of ray casts for randomly generated states.
